% ============================================================
% APPENDIX
% ============================================================
\appendix

% ============================================================
% A. EXTENDED RELATED WORK
% ============================================================
\section{Extended Related Work}
\label{app:related}

\subsection{Adaptive Compute: Reasoning vs.\ Agent Settings}

The adaptive test-time compute literature spans two distinct
settings. In the \emph{reasoning} setting, methods optimize
single-turn chain-of-thought generation on homogeneous benchmarks
(MATH, GSM8K, GPQA).
Signal-based approaches use token-level confidence or entropy to
decide when to extend
thinking~\cite{seag2025, cats2025, corefine2026,
thinkjustenough2025, s1_2025, han2025tokenbudget}.
RL-based methods learn think/no-think policies through reinforcement
learning~\cite{adaptthink2025, thinkless2025, aggarwal2025l1}.
Hidden-state probing methods train lightweight classifiers on
internal representations to estimate task
difficulty~\cite{diffadapt2025, zhu2025llmalreadyknows}.
Because reasoning benchmarks share similar task semantics, signal
direction happens to be consistent across tasks, and the
fixed-direction assumption is never tested.

In the \emph{agent} setting, methods operate in multi-step
interactive environments with external optimizers. Vote-based
methods~\cite{catts2026} sample multiple candidate actions and
trigger evaluation when disagreement is high.
\citet{paglieri2025learning} learn when to invoke a planner at
the episode level. Agentic RL methods~\cite{arpo2025} use
entropy-based adaptive rollout at tool-call steps. All agent-setting
methods inherit the fixed-direction assumption from the reasoning
literature without verifying it. Our work is the first to
systematically study signal semantics across heterogeneous agent
environments, where we find that the direction reverses.

\subsection{Detailed Baseline Comparison}
\label{app:method-comparison}

Table~\ref{tab:baseline-detail} provides a detailed comparison of all
baselines evaluated in this paper. All six methods share a common
assumption: a fixed mapping from signal value to compute need.

\begin{table*}[h]
\caption{Detailed comparison of baselines. All methods assume a
fixed signal--utility direction. $^\dagger$Requires a separate labeled dataset for calibration.}
\label{tab:baseline-detail}
\centering\small
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\textwidth}{l W W N W}
\toprule
Method & Signal & Direction & Granularity & Mechanism \\
\midrule
CATTS~\cite{catts2026} & Vote entropy + margin & High disagreement $\to$ trigger
  & Per-step & $K{=}5$ forward passes, arbiter on disagreement \\
SEAG$^\dagger$~\cite{seag2025} & Mean token confidence & Low conf.\ $\to$ search
  & Per-problem & Platt scaling, search depth allocation \\
CaTS$^\dagger$~\cite{cats2025} & Calibrated confidence & Low conf.\ $\to$ generate
  & Per-attempt & Self-calibration, Best-of-N early stopping \\
CoRefine$^\dagger$~\cite{corefine2026} & Conv1D confidence & Low conf.\ $\to$ refine
  & Per-problem & Halt / re-examine / redirect controller \\
s1\_budget~\cite{s1_2025} & Token budget & Fixed budget $\to$ stop
  & Per-problem & Budget-constrained generation \\
AUQ~\cite{auq2026} & Verbalized uncertainty & High uncert.\ $\to$ reflect
  & Per-step & Uncertainty-aware memory + reflection \\
\midrule
\textbf{EAAG} & \textbf{Multi (auto)} & \textbf{Learned}
  & \textbf{Per-step} & Explore + reason + LASSO direction learning \\
\bottomrule
\end{tabularx}
\vspace{2pt}\\
\end{table*}
\subsection{Adaptive Computation in Neural Networks}

The idea of adapting computation to input difficulty predates LLMs.
Adaptive Computation Time~\cite{graves2016act} allows RNNs to learn
how many computational steps to take per input. Early exit
methods~\cite{teerapittayanon2016branchynet, elhoushi2024layerskip}
attach classifiers at intermediate transformer layers and exit early
for easy inputs. Mixture-of-Experts
architectures~\cite{shazeer2017moe, raposo2024mixtureofdepths}
activate only a subset of parameters per token, adapting capacity
rather than depth. These methods adapt computation within a single
forward pass. Our work operates at a different granularity: we
adapt whether to invoke an external optimizer at each decision step,
and the key challenge is not input difficulty but the
\emph{direction} of the signal--utility relationship.

\subsection{LLM Confidence Estimation and Calibration}
\label{app:calibration}

\citet{kadavath2022knowwhat} show that larger models are
well-calibrated on multiple-choice tasks and can estimate P(True)
for their own outputs.
\citet{tao2025revisiting} and \citet{heo2025llmuncertainty} find
that calibration quality varies substantially across task types,
providing converging evidence that uncertainty semantics are
domain-dependent.
\citet{yadkori2024believe} propose decoupling epistemic and
aleatoric uncertainty in LLM outputs, a distinction conceptually
related to our Type~I/Type~D decomposition.
Our work extends these findings: not only does calibration
\emph{quality} vary across tasks, but the \emph{direction} of the
uncertainty--utility relationship reverses entirely.

\subsection{Metareasoning and Value of Computation}

Rational metareasoning~\cite{russell1991right,
horvitz1987reasoning} formalizes the value of computation (VOC)
as a guide for allocating cognitive resources. The classical
framework assumes VOC~$\geq 0$ because an agent can always ignore
computation results. \citet{lieder2020resource} generalize this to
resource-rational analysis, modeling cognition as optimal use of
limited computational resources.
\citet{desabbata2024metareasoning} apply metareasoning to LLMs,
analyzing when to extend chain-of-thought reasoning.
Our work extends this framework by showing that VOC can be
\emph{negative} in the agent setting when the evaluator and executor
are the same model. In this regime, wrong-direction gating does not
merely waste computation (VOC~$= 0$) but actively degrades
performance (VOC~$\ll 0$), violating the classical assumption.
Direction discovery is therefore a necessary precondition for
non-negative VOC across environments.

\subsection{Simpson's Paradox in Machine Learning}
\label{app:simpson}

Simpson's paradox~\cite{simpson1951interpretation,
DBLP:books/acm/22/Pearl22n} occurs when a trend present in subgroups
reverses upon aggregation. In machine learning, this phenomenon has
been documented in fairness~\cite{kdd2023simpson}, causal
inference, and treatment effect estimation. Our Two-Source Model
identifies a new instance: within Type~I states, entropy negatively
predicts utility; within Type~D states, entropy positively predicts
utility; but the marginal correlation depends on the mixture
proportion $p_I(\mathcal{E})$, which varies across environments.
This makes direction reversal a structural consequence of
aggregating heterogeneous uncertainty sources.

\subsection{Concurrent Work Statement}

Several 2025--2026 papers independently explore adaptive compute.
CATTS~\cite{catts2026} (Feb 2026) and CoRefine~\cite{corefine2026}
(Feb 2026) are concurrent with no methodological overlap.
AdaptThink~\cite{adaptthink2025} (May 2025) and
Thinkless~\cite{thinkless2025} (May 2025) predate our work but
operate in the reasoning setting. DiffAdapt~\cite{diffadapt2025}
(Oct 2025) assumes a universal U-shaped entropy pattern that our
cross-environment evidence disproves.
AUQ~\cite{auq2026} (Jan 2026) uses verbalized uncertainty for
agent reflection but assumes fixed direction.
Our distinct contribution is threefold: first to (a)~demonstrate
direction reversal across environments and backbones, (b)~formalize
the mechanism via a Two-Source Model grounded in Simpson's paradox,
and (c)~prove that direction discovery is a necessary condition for
cross-environment non-negative VOC.


% ============================================================
% B. THEORETICAL MATERIAL
% ============================================================
\section{Theoretical Material}
\label{app:theory}

This appendix collects all theoretical material referenced in
Section~\ref{sec:signal-landscape}: the proof of
Proposition~\ref{prop:necessity}, the wrong-direction damage
quantification, the VOC negativity analysis, and the full derivation
and per-environment mapping of the Two-Source Model.

\subsection{Proof of Proposition~\ref{prop:necessity}
  (Necessity of Direction Discovery)}
\label{app:proof-necessity}

\begin{proposition*}[Restated]
Suppose there exist two environments $\mathcal{E}_1, \mathcal{E}_2$
with opposite true directions: $d^*(\mathcal{E}_1) = +1$ and
$d^*(\mathcal{E}_2) = -1$. Then no fixed-direction gate $g_d$ can
simultaneously satisfy
$\mathrm{SR}(g_d, \mathcal{E}_1) \geq \mathrm{SR}(\mathrm{base},
\mathcal{E}_1)$ and
$\mathrm{SR}(g_d, \mathcal{E}_2) \geq \mathrm{SR}(\mathrm{base},
\mathcal{E}_2)$.
\end{proposition*}

\begin{proof}
Define a fixed-direction gate as
$g_d(s) = \mathbf{1}[d \cdot \sigma(s) > \theta]$
for $d \in \{+1, -1\}$ and threshold $\theta$.

Define $d^*(\mathcal{E}) = +1$ if
$\mathrm{Corr}(\sigma(s), U(T,s)) > 0$ in $\mathcal{E}$ (Type~D),
and $d^*(\mathcal{E}) = -1$ if
$\mathrm{Corr}(\sigma(s), U(T,s)) < 0$ (Type~I).

\textbf{Lemma (Wrong-direction harm).} If $d \neq d^*(\mathcal{E})$,
then
$\mathrm{SR}(g_d, \mathcal{E}) \leq
 \mathrm{SR}(\mathrm{base}, \mathcal{E}) - \delta(\mathcal{E})$
where $\delta(\mathcal{E}) > 0$ depends on signal strength
$|\rho(\mathcal{E})|$.

\emph{Proof of Lemma}: By definition of $d^*$, states with high
$d \cdot \sigma(s)$ are states where $U < 0$. The gate triggers $T$
at these states, causing the agent to adopt worse actions. The base
policy, which never triggers $T$, avoids this systematic harm.

\textbf{Main argument:}

\emph{Case 1}: $d = +1$. In $\mathcal{E}_2$ where $d^* = -1$,
$d \neq d^*(\mathcal{E}_2)$. By Lemma:
$\mathrm{SR}(g_{+1}, \mathcal{E}_2) \leq
 \mathrm{SR}(\mathrm{base}, \mathcal{E}_2) - \delta(\mathcal{E}_2)
 < \mathrm{SR}(\mathrm{base}, \mathcal{E}_2)$.
Constraint violated.

\emph{Case 2}: $d = -1$. In $\mathcal{E}_1$ where $d^* = +1$,
by symmetric argument, constraint violated.

Since $d \in \{+1, -1\}$ and both cases lead to violation, no fixed
$d$ satisfies both constraints simultaneously.
\end{proof}

\subsection{Wrong-Direction Damage Quantification}
\label{app:wrong-direction}

The proof above bounds wrong-direction harm by $\delta(\mathcal{E})$
but does not specify its magnitude. Table~\ref{tab:wrong-direction}
reports the empirically observed damage from the BSW ablation,
showing that $\delta(\mathcal{E})$ scales with signal strength
$|\rho(\mathcal{E})|$.

\begin{table}[h]
\centering\small
\caption{Wrong-direction damage scales with signal strength.}
\label{tab:wrong-direction}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lcccc}
\toprule
Environment & Correct SR & Wrong SR & $\Delta$ SR & $|\rho|$ (strongest) \\
\midrule
HotpotQA   & 95.2\% & 58.2\% & $-$37.0 & 0.494 \\
WebShop    & 43.8\% & 21.4\% & $-$22.4 & 0.444 \\
FEVER      & 49.8\% & 13.0\% & $-$36.8 & 0.619 \\
APPS Intro & 66.0\% & 62.5\% & $-$3.5  & 0.155 \\
\bottomrule
\end{tabular}%
}
\end{table}

Environments with stronger signal--utility coupling ($|\rho|$)
suffer larger SR degradation when direction is reversed
($R^2 > 0.5$). On HotpotQA, the MLP gate with wrong direction
achieves 45.3\%, \emph{below} the 49.0\% base, demonstrating that
a more powerful gate with wrong direction is \emph{worse} than a
simpler gate with wrong direction (logistic: 62.0\%), because it
more precisely targets harmful states.

\subsection{VOC Non-Negativity Scope}
\label{app:voc}

In the classical VOC framework~\cite{russell1991right},
$\mathrm{VOC}(T, s) =
 \mathbb{E}[\max(V(a_T), V(a_{\mathrm{base}}))]
 - V(a_{\mathrm{base}}) \geq 0$
because the agent can \emph{disregard} the computation result if
unfavorable. Our setting violates this: when the agent uses an
evaluator/optimizer $T$, the evaluator's assessment \emph{is} the
agent's decision. The agent cannot separately evaluate the
evaluator's output because the evaluator \emph{is} the evaluation
mechanism.

Under evaluator-executor identity, VOC can be negative:
$\mathrm{VOC}(T, s) =
 \mathbb{E}[V(a_T(s))] - V(a_{\mathrm{base}}(s))$.
If $T$ systematically selects worse actions in Type~I states,
$\mathrm{VOC} < 0$ for those states. This explains why
wrong-direction gating is not merely wasteful
($\mathrm{VOC} = 0$) but actively harmful
($\mathrm{VOC} < 0$).

\subsection{Two-Source Model: Full Derivation}
\label{app:model-derivation}

At each step $t$, state $s_t$ belongs to one of two latent types:
$s_t \in \text{Type~I}$ with probability $p_I(\mathcal{E})$,
$s_t \in \text{Type~D}$ with probability $1 - p_I(\mathcal{E})$.

Conditional utility models:
\begin{align}
  U(T, s) \mid s \in \text{Type~I} &= -\alpha \cdot H(s) + \varepsilon_I, \quad \alpha > 0 \\
  U(T, s) \mid s \in \text{Type~D} &= +\beta \cdot H(s) + \varepsilon_D, \quad \beta > 0
\end{align}

The marginal correlation:
\begin{align}
  \mathrm{sign}(\rho(\mathcal{E})) &= \mathrm{sign}(\beta \cdot (1 - p_I) - \alpha \cdot p_I) \\
  &= \mathrm{sign}(\beta - (\alpha + \beta) \cdot p_I)
\end{align}

Reversal threshold: $p_I^* = \beta / (\alpha + \beta)$.

\subsection{Per-Environment Type Mapping}
\label{app:env-mapping}

Table~\ref{tab:env-type-mapping} maps each evaluation environment
to its dominant uncertainty type, the direction of $\rho$ predicted
by Eq.~\eqref{eq:direction}, and the observed $\rho$ on Qwen3-4B.
Multi-backbone $\rho$ values are reported in
Table~\ref{tab:signal-discovery} of the main text and in
Appendix~\ref{app:multi-backbone}. The mapping is primarily
determined by the environment's information structure, but the
model backbone modulates the effective $p_I$: the same environment
can shift across the threshold $p_I^*$ depending on which model
processes it (e.g., FEVER: $\rho{=}{-}0.156$ on Phi-3.5 vs.\
$\rho{=}{+}0.428$ on Llama-3.1). Structural signals such as
\texttt{step\_count} are more robust across backbones than entropy,
which is why EAAG cannot rely on environment-level heuristics
alone and must discover direction per (environment, backbone)
pair from interaction data.

\begin{table*}[h]
\caption{Environment-to-type mapping. Predicted $\rho$ direction
follows from Eq.~\eqref{eq:direction}; observed values are from
Qwen3-4B.}
\label{tab:env-type-mapping}
\centering\small
\resizebox{\textwidth}{!}{%
\begin{tabular}{llccl}
\toprule
Environment & Dominant Type & Pred.\ $\rho$ & Obs.\ $\rho$ & Rationale \\
\midrule
HotpotQA       & Type~I (info-poverty) & $-$ & $-0.041$ & Requires retrieving specific evidence \\
FEVER          & Type~I (info-poverty) & $-$ & $-0.119$ & Depends on finding supporting facts \\
TWExpress      & Type~I (info-poverty) & $-$ & $-0.290$ & Sparse reward, long info-gathering \\
\midrule
APPS Intv.     & Type~D (decision-diff) & $+$ & $+0.317$ & Multiple valid solution strategies \\
\midrule
APPS Intro     & Mixed ($p_I \approx p_I^*$) & $\approx 0$ & $+0.012$ & Mix of info-gathering \& code design \\
WebShop        & Mixed (entropy weak) & $\approx 0$ & $-0.019$ & Choice-dominated but entropy weak \\
\midrule
CRUXEval       & Weak signal & $\approx 0$ & $-0.048$ & Short trajectories, little variation \\
Plancraft      & Weak signal & $\approx 0$ & $-0.016$ & Rollouts harmful; entropy uninformative \\
\bottomrule
\end{tabular}%
}
\end{table*}

The mapping confirms the model's core prediction: the sign of the
entropy--utility correlation is determined by which uncertainty
source dominates, not by properties of the entropy signal itself.

\subsection{Position on the $p_I$ Axis}
\label{app:pi-axis}

{\color{red} \textbf{[FIGURE NEEDED: Figure A1.]}
\textbf{$p_I$ axis diagram: visual summary of the Two-Source Model.}

\textbf{Layout:} Single horizontal axis, full column width.

\textbf{Data:} From Table~\ref{tab:env-type-mapping} above.
8 environments with estimated $p_I$ values and observed $\rho$.

\textbf{Visual design:}
\begin{itemize}[nosep]
\item Horizontal axis from $p_I = 0$ (left, ``Pure Type~D'')
  to $p_I = 1$ (right, ``Pure Type~I'')
\item Vertical dashed line at $p_I^* = \beta/(\alpha+\beta)$
  labeled ``Reversal threshold $p_I^*$''
\item Left region (green background): positive $\rho$, trigger
  on high uncertainty is correct
\item Right region (red background): negative $\rho$, trigger
  on high uncertainty is harmful
\item Each environment = a labeled dot on the axis at its
  estimated $p_I$
\item Dot size proportional to $|\rho|$ (larger = stronger signal)
\item Dot color: green for $\rho > 0$, red for $\rho < 0$,
  gray for $|\rho| < 0.05$
\item Label each dot with environment name + $\rho$ value
\item Below axis: bracket grouping ``Type~D dominant'' (left),
  ``Mixed / Weak'' (center), ``Type~I dominant'' (right)
\end{itemize}

\textbf{Style:} TikZ, clean and publication-quality.
Similar to a number-line diagram commonly used in physics papers.

\textbf{Key message:} The 8 environments span the full $p_I$ axis;
no single fixed direction works for all environments.}

Estimated positions on the $p_I$ axis (0 = pure Type~D, 1 = pure
Type~I):
\begin{itemize}[nosep,leftmargin=*]
\item $p_I \approx 0.2$: APPS Interview (strong positive $\rho = +0.317$)
\item $p_I \approx 0.45$: APPS Intro (near-zero $\rho = +0.012$)
\item $p_I \approx 0.5$: WebShop (entropy $\rho \approx 0$, but non-entropy signals work)
\item $p_I \approx 0.55$: CRUXEval (weak negative $\rho = -0.048$)
\item $p_I \approx 0.6$: Plancraft (weak negative, rollouts harmful)
\item $p_I \approx 0.7$: HotpotQA (negative $\rho = -0.041$)
\item $p_I \approx 0.8$: TWExpress (negative $\rho = -0.290$)
\item $p_I \approx 0.9$: FEVER (strong negative $\rho = -0.119$)
\end{itemize}


% ============================================================
% C. MULTI-BACKBONE VERIFICATION
% ============================================================
\section{Multi-Backbone Verification}
\label{app:multi-backbone}

We validate our core findings on three backbones from three vendors:
Qwen3-4B-Instruct (Alibaba, 4B), Phi-3.5-mini-instruct (Microsoft,
3.8B), and Llama-3.1-8B-Instruct (Meta, 8B) across all 8
environments.

\subsection{Signal Direction: Environment $\times$ Model Interaction}

The multi-backbone results
(Table~\ref{tab:signal-discovery} in main text) reveal that 4/5
environments with $\geq$2 significant signals show cross-model
direction variation. Structural signals (\texttt{step\_count}) are
model-invariant while entropy signals are model-dependent.

{\color{blue} \textbf{[DATA NEEDED: Table A6.]} Full multi-backbone results.
8 environments $\times$ 3 backbones $\times$ multiple signals.
Extends Table 1 in main text with all signal types, not just entropy.
Shows that step\_count direction is stable while entropy flips.}

\subsection{Fixed-Direction Baselines Fail Across Backbones}

On Phi-3.5 APPS: all 6 CB methods score \emph{below} base\_only
(59\% $\to$ 27--36\%), demonstrating that fixed-direction assumption
is catastrophically wrong when the model changes.

{\color{blue} \textbf{[DATA NEEDED: Table A7.]} Cross-backbone baseline results.
Methods $\times$ environments for Phi-3.5 and Llama-3.1 backbones.
Shows that the same baselines that work on Qwen3 fail on other backbones,
while EAAG adapts successfully.}

{\color{red} \textbf{[FIGURE NEEDED: Figure A4.]}
\textbf{Multi-backbone signal direction heatmap.}

\textbf{Layout:} Heatmap, 3 panels side by side (one per backbone).
Each panel: rows = 8 environments, columns = key signals
(step\_count, token\_entropy, evidence\_count, etc.).
Cell color = Spearman $\rho$ (red-white-blue diverging colormap).

\textbf{Data:} From \texttt{tab\_signal\_discovery/} CSV files,
extended to all 3 backbones.

\textbf{Visual design (extends main-text fig1\_signal\_heatmap):}
\begin{itemize}[nosep]
\item Same color scale across all 3 panels for comparison
\item Annotate $\rho$ values inside cells (like current
  fig1\_signal\_heatmap.png)
\item Bold border on cells where sign flips across backbones
\item Panel titles: ``Qwen3-4B'', ``Phi-3.5'', ``Llama-3.1-8B''
\end{itemize}

\textbf{Key message:} Entropy direction flips across backbones
in 4/5 environments with significant signal. Structural signals
(step\_count) are model-invariant. Justifies per-(env, model)
direction discovery.

\textbf{Data source:} Run signal discovery (Spearman $\rho$) on
exploration data from each backbone. Fields: env, backbone, signal,
$\rho$, $p$-value.}

{\color{red} \textbf{[FIGURE NEEDED: Figure A5.]}
\textbf{Cross-backbone SR comparison: EAAG vs.\ best fixed baseline.}

\textbf{Layout:} Grouped bar chart.
x-axis = 8 environments. 2 groups of 3 bars each:
Group 1 (blue shades): best fixed-direction baseline SR on
Qwen3 / Phi-3.5 / Llama-3.1.
Group 2 (red shades): EAAG SR on Qwen3 / Phi-3.5 / Llama-3.1.

\textbf{Data:} From Table A7.

\textbf{Visual design:}
\begin{itemize}[nosep]
\item Blue bars = fixed baseline, Red bars = EAAG
\item Lighter shade = Qwen3, Medium = Phi-3.5, Darker = Llama
\item Annotate $\Delta$SR between EAAG and fixed baseline for
  each backbone
\item Sort environments by magnitude of cross-backbone variation
\end{itemize}

\textbf{Key message:} Fixed baselines are brittle across backbones
(large SR variance); EAAG is robust (small SR variance).}


% ============================================================
% D. SIGNAL AND FEATURE ANALYSES
% ============================================================
\section{Signal and Feature Analyses}
\label{app:signal-analyses}

This appendix collects all empirical analyses of signal identity,
signal sufficiency, and the Two-Source Model's predictions
referenced in Section~\ref{sec:empirical-landscape} and
Section~\ref{sec:theory-verification}. These are diagnostic results
about \emph{which signals carry information and how}, kept separate
from the experimental setup details in
Appendix~\ref{app:experiments}.

\subsection{Per-Environment Feature Selection}
\label{app:feature-selection}

This subsection expands on the signal-identity finding referenced
in Section~\ref{sec:empirical-landscape}: that the most informative
signal varies across environments, with no single feature
dominating universally.

Sparse feature selection across environments reveals that different
environments rely on entirely different signal subsets.
\texttt{step\_count} is informative in 6 of 7 environments where
any signal is informative, suggesting that trajectory progress is
a broadly useful proxy for information accumulation. WebShop is
the notable exception: its dominant signals are
\texttt{num\_available\_actions} and task-specific features such
as \texttt{price\_mentioned} and \texttt{action\_is\_click},
reflecting its choice-dominated structure where decision-space
size matters more than information accumulation. Token entropy,
despite being the standard signal in prior adaptive-compute work,
is selected in fewer than half of environments. No single signal
is universally informative, motivating the multi-signal design of
EAAG.

{\color{blue} \textbf{[DATA NEEDED: Table A1.]} Environment $\times$ Feature
matrix showing LASSO coefficients. Rows = 8 environments.
Columns = step\_count, token\_entropy, state\_length, action\_entropy,
num\_available\_actions, has\_output, + LLM-generated features per env.
Values = LASSO coefficient (0 = not selected, sign = direction).
This is the detailed version of the heatmap referenced in
Section~\ref{sec:empirical-landscape}.}

\subsection{LLM-Generated Features}
\label{app:llm-features}

{\color{blue} \textbf{[DATA NEEDED: Table A2.]} LLM-generated features per
environment. Columns: Environment, LLM Prompt (abbreviated), Generated
Features, LASSO Selected (Y/N), Coefficient. Shows the LLM reasoning
step's concrete output for each environment.}

\subsection{AUC Hierarchy: Single vs.\ Multi-Signal Gates}
\label{app:auc}

This subsection expands on the claim in
Section~\ref{sec:empirical-landscape} that even the best single
signal is insufficient for predicting rollout value. We compare
three gate families on held-out data across four representative
environments (HotpotQA, APPS Intro, WebShop, FEVER):
(i) a single-signal gate using token entropy, (ii) a multi-signal
logistic gate over the LASSO-selected feature set, and
(iii) a hidden-state probe trained on the model's final-layer
representations.

A clear hierarchy emerges. Single token entropy achieves
AUC${\approx}$0.50, indistinguishable from chance, confirming that
no single scalar reliably predicts $U(T,s)$. Multi-signal logistic
gates reach AUC${\approx}$0.83, a substantial jump achieved purely
by combining cheap, interpretable features. Hidden-state probes
push this further to AUC${\approx}$0.90, but at the cost of opacity
and per-step inference overhead. The 0.50~$\to$~0.83 gap is the
relevant one for our method design: it shows that the bottleneck
is signal \emph{combination}, not gate \emph{capacity}, which is
why EAAG invests its complexity in feature discovery rather than
in a deeper gate architecture (cf.\ the gate-complexity ablation
in Table~\ref{tab:complexity-ablation}).

{\color{blue} \textbf{[DATA NEEDED: Table A0.]} AUC table.
Rows = 4 environments (HotpotQA, APPS Intro, WebShop, FEVER).
Columns = single-entropy AUC, multi-signal logistic AUC,
hidden-state probe AUC.}

{\color{red} \textbf{[FIGURE NEEDED: Figure A3.]}
\textbf{AUC hierarchy bar chart.}

\textbf{Layout:} Grouped bar chart, 1 panel.
x-axis = 4 environments. 3 bars per env: single-entropy,
multi-signal logistic, hidden-state probe.
y-axis = AUC.

\textbf{Data:} Same as Table A0.

\textbf{Visual design:}
\begin{itemize}[nosep]
\item 3 colors: light gray (single), blue (multi-signal),
  dark blue (hidden-state)
\item Horizontal dashed line at AUC = 0.5 (chance level)
\item Annotate numbers on top of each bar
\item Annotate the 0.50 $\to$ 0.83 gap with a bracket
  labeled ``signal combination gain''
\end{itemize}

\textbf{Key message:} Single entropy $\approx$ chance; multi-signal
logistic achieves 83\%; hidden-state probe adds only marginal gain.
Justifies EAAG's multi-signal + simple-gate design.}

\subsection{Temporal Dynamics (Prediction P1)}
\label{app:temporal}

{\color{blue} \textbf{[DATA NEEDED: Table A4.]} P1 Temporal Dynamics full data.
Columns: Environment, $\rho_{\text{early}}$ (step $\leq$ median),
$\rho_{\text{late}}$ (step $>$ median), $p_{\text{early}}$,
$p_{\text{late}}$. All 8 environments. Shows the temporal shift
pattern for each.}

\subsection{Signal Identity Alignment (Prediction P3)}
\label{app:signal-alignment}

{\color{blue} \textbf{[DATA NEEDED: Table A5.]} P3 Signal Identity Alignment.
Columns: Environment, Dominant Type, Strongest Signal, $|\rho|$,
Signal Interpretation. Shows what each environment's strongest
signal measures and how it aligns with the predicted type.}

\subsection{Controlled Information Manipulation}
\label{app:controlled}

To provide stronger evidence that information structure drives
signal semantics, we manipulate information availability within
HotpotQA itself. \emph{InfoPoor} (search returns only the first
sentence) reduces base SR to 44.0\% while \emph{InfoRich} (gold
evidence injected at start) raises it to 82.0\%. The signal
hierarchy shifts: in InfoPoor, \texttt{step\_count} dominates
($\rho{=}{-}0.608$) while entropy carries a weak positive signal
($\rho{=}{+}0.119$, $n{=}317$; the Two-Source Model predicts
entropy should be negative in info-poor settings, but the small
sample and mixed early-step Type~D component may explain the
weakly positive sign; the key finding is that \texttt{step\_count}
overtakes entropy as the dominant signal); in InfoRich, entropy
becomes the dominant positive signal ($\rho{=}{+}0.311$) while
\texttt{step\_count} weakens ($\rho{=}{-}0.147$). This
demonstrates that the \emph{identity} of the most informative
signal, not just its direction, is a function of the environment's
information structure.

{\color{blue} \textbf{[DATA NEEDED: Table A9.]} InfoPoor vs InfoRich
full signal correlation table for all features. Columns: Condition,
Base SR, $\rho$ for each feature, strongest signal.}

{\color{red} \textbf{[FIGURE NEEDED: Figure A6.]}
\textbf{Controlled information manipulation: signal hierarchy shift.}

\textbf{Layout:} $1 \times 2$ panels.

\textbf{Panel (a): Signal $\rho$ comparison.}
\begin{itemize}[nosep]
\item Paired bar chart: x-axis = signals (step\_count, entropy,
  evidence\_count, etc.)
\item Two bars per signal: InfoPoor (dark blue) vs.\ InfoRich (light green)
\item y-axis = Spearman $\rho$
\item Horizontal dashed line at $\rho = 0$
\item Highlight the signal that ``wins'' in each condition
  (step\_count for InfoPoor, entropy for InfoRich)
\end{itemize}

\textbf{Panel (b): EAAG gate weight comparison.}
\begin{itemize}[nosep]
\item Same layout as (a) but y-axis = LASSO coefficient $w_i$
\item Shows that EAAG automatically shifts which signal it
  relies on based on the information structure
\end{itemize}

\textbf{Data:} From controlled HotpotQA experiments.
Run exploration + LASSO separately for InfoPoor and InfoRich
conditions.

\textbf{Key message:} When information availability changes,
the identity of the dominant signal shifts, and EAAG's LASSO
automatically tracks this shift. This is the strongest causal
evidence that information structure drives signal semantics.}


% ============================================================
% E. EXPERIMENTAL DETAILS
% ============================================================
\section{Experimental Details}
\label{app:experiments}

\subsection{Environment Specifications}

For each of the 8 environments, we provide:
\begin{itemize}[nosep,leftmargin=*]
\item Task description and action space
\item State representation and reward function
\item Episode length distribution
\item Optimizer $T$ implementation details:
  \begin{itemize}[nosep]
  \item HotpotQA, FEVER, TWExpress, Plancraft: per-action evaluation
    (generate $K{=}5$ candidates, evaluate each, select best)
  \item APPS Intro, APPS Interview, CRUXEval: $K$-variant sampling
    (generate $K{=}3$ code solutions, test each, select passing one)
  \item WebShop: LLM-Propose-$K$ (LLM proposes $K{=}3$ action
    sequences, simulate each, select highest-reward)
  \end{itemize}
\end{itemize}

\subsection{EAAG Implementation Details}
\label{app:eaag-details}

This appendix expands on the implementation details deferred from
Section~\ref{sec:learn}. Algorithm~\ref{alg:eaag} gives the full
pipeline; the paragraphs that follow describe each component.

\begin{algorithm}[h]
\caption{EAAG: Environment-Aware Adaptive Gating}
\label{alg:eaag}
\begin{algorithmic}[1]
\Require Environment $\mathcal{E}$, agent policy $\pi$, optimizer
         $T$, universal feature set $\phi_{\text{univ}}$, LLM
         $\mathcal{M}$, exploration budget $N_{\text{explore}}$,
         exploration rate $\varepsilon_{\text{explore}}$,
         deployment decay schedule $\varepsilon_{\text{deploy}}(t)$,
         retrain interval $k$, gate threshold $\tau$
\Ensure Deployed gate $g: \mathcal{S} \to \{0, 1\}$

\Statex \textbf{// Phase 1: Explore}
\State $\mathcal{D} \gets \emptyset$
\For{episode $= 1, \dots, N_{\text{explore}}$}
  \State Run $\pi$ in $\mathcal{E}$; at each step $t$, invoke $T$
         with probability $\varepsilon_{\text{explore}}$
         independent of state
  \State Record $(s_t, U_t)$ pairs into $\mathcal{D}$
\EndFor

\Statex \textbf{// Phase 2: Reason}
\State $(p_{\mathcal{E}}, \mathcal{H}) \gets
       \mathcal{M}(\mathrm{summary}(\mathcal{D}))$
       \Comment{environment profile + feature hypotheses}
\State $\phi_{\text{cand}} \gets \phi_{\text{univ}} \cup
       \mathrm{extract}(\mathcal{H})$
\State Recompute $\phi_{\text{cand}}(s)$ for all
       $s \in \mathcal{D}$

\Statex \textbf{// Phase 3: Learn}
\State Standardize $\phi_{\text{cand}}$ on $\mathcal{D}$
\State Select regularization strength $C$ via 5-fold CV on
       held-out log-loss
\State Fit $\ell_1$-regularized logistic regression:
       $\mathbf{w}, b \gets \arg\min\,
       \mathcal{L}_{\text{logistic}}(\mathbf{w}, b; \mathcal{D}) +
       \tfrac{1}{C}\|\mathbf{w}\|_1$
\State $\phi \gets$ features with non-zero $w_i$
\State $g(s) \gets
       \mathbf{1}[\sigma(\mathbf{w}^\top \phi(s) + b) > \tau]$

\Statex \textbf{// Phase 4 (optional): Online adaptation}
\For{deployment episode $t = 1, 2, \dots$}
  \State Deploy $g$; with probability
         $\varepsilon_{\text{deploy}}(t)$ override with random
         trigger; append new $(s, U)$ pairs to $\mathcal{D}$
  \If{$t \bmod k = 0$}
    \State Refit $g$ on updated $\mathcal{D}$ (repeat Phase 3,
           reusing $\phi_{\text{cand}}$)
  \EndIf
\EndFor

\State \Return $g$
\end{algorithmic}
\end{algorithm}

\paragraph{Exploration budget.}
Phase 1 runs for $N_{\text{explore}} = 50$ episodes with
$\varepsilon_{\text{explore}} = 0.5$. Sensitivity to
$N_{\text{explore}}$ is reported in
Appendix~\ref{app:additional}; SR is stable for
$N_{\text{explore}} \geq 30$.

\paragraph{Feature standardization.}
Before fitting in Phase 3, all features in $\phi_{\text{cand}}$
are standardized to zero mean and unit variance on $\mathcal{D}$.
This ensures that the interpretable weight signs discussed in
Section~\ref{sec:learn} are not confounded by scale differences
across heterogeneous features (e.g., \texttt{step\_count} and
\texttt{token\_entropy} live on very different scales).

\paragraph{Sparse solver.}
We induce sparsity using $\ell_1$-regularized logistic regression
(\texttt{scikit-learn}'s \texttt{LogisticRegression} with
\texttt{liblinear} solver). The regularization strength $C$ is
selected by 5-fold cross-validation over the grid
$C \in \{0.01, 0.03, 0.1, 0.3, 1.0, 3.0, 10.0\}$, optimizing
held-out log-loss on $\mathcal{D}$. Features with zero coefficient
after fitting are dropped from the deployed gate.

\paragraph{Gate threshold.}
The decision threshold $\tau$ in Eq.~\eqref{eq:gate} is fixed to
$0.5$ for all environments. Sensitivity across
$\tau \in [0.3, 0.7]$ is reported in
Appendix~\ref{app:additional}.

\paragraph{Online adaptation.}
During deployment, $\varepsilon_{\text{deploy}}$ decays from $0.1$
to $0$ over the first 100 episodes, and the gate is refit every
$k = 30$ episodes using the accumulated $\varepsilon$-greedy
samples. Refitting reuses the same solver configuration as the
initial training, so no additional hyperparameter tuning is
required at deployment time.

\subsection{Computational Cost Breakdown}

{\color{blue} \textbf{[DATA NEEDED: Table A3.]} Method $\times$ Cost component.
Columns: Method, Phase-1 Episodes, Phase-1 Compute (GPU-hrs),
Per-Step Overhead (ms), Gate Training Time, Total Cost for 500 Episodes.
Shows EAAG's cost advantage: 50 exploration episodes, $\sim$2 GPU-hrs,
0ms gate overhead, $<$1s training.}

{\color{red} \textbf{[FIGURE NEEDED: Figure A7.]}
\textbf{Full Pareto frontier plot (all 8 environments).}

\textbf{Layout:} $2 \times 4$ grid (all 8 environments).
Same format as main-text Pareto plot (Figure~\ref{fig:pareto})
but including APPS Interview and CRUXEval.

\textbf{Data:} Same as main-text Pareto plot but complete.

\textbf{Visual design:} Identical to main-text version:
star for EAAG, triangle for baselines, circle for bounds,
Pareto frontier dashed line. This is the ``full version'' of the
main-text figure which only shows 6 representative environments.

\textbf{Key message:} Pareto dominance holds across all 8 environments,
including the 2 not shown in main text.}

{\color{red} \textbf{[FIGURE NEEDED: Figure A8.]}
\textbf{Computational cost breakdown}
(ref: CoRefine Table~2 token efficiency format).

\textbf{Layout:} Stacked bar chart.
x-axis = methods (EAAG, CaTS, SEAG, CoRefine, CATTS, AUQ, s1).
y-axis = total compute (GPU-hours for 500 episodes).

\textbf{Stacked components:}
\begin{itemize}[nosep]
\item Dark color: calibration / exploration overhead
\item Light color: deployment compute (rollouts)
\end{itemize}

\textbf{Data:} From Table A3. Compute total cost as
$\text{overhead} + \text{deployment cost per episode} \times 500$.

\textbf{Visual design:}
\begin{itemize}[nosep]
\item Stacked bars with two components
\item Annotate total on top of each bar
\item EAAG highlighted with red border
\item Sort by total cost (ascending)
\end{itemize}

\textbf{Key message:} EAAG's 50-episode exploration overhead is
comparable to or less than calibration overhead of other methods,
and deployment cost is lower due to selective triggering.}


% ============================================================
% F. ADDITIONAL ANALYSES
% ============================================================
\section{Additional Analyses}
\label{app:additional}

\subsection{Hyperparameter Sensitivity}

{\color{red} \textbf{[FIGURE NEEDED: Figure A2.]}
\textbf{Hyperparameter sensitivity plots.}

\textbf{Layout:} $1 \times 2$ panels side by side.

\textbf{Panel (a): Exploration budget sensitivity.}
\begin{itemize}[nosep]
\item x-axis: $N_{\text{explore}} \in \{10, 20, 30, 50, 100\}$
\item y-axis: SR (\%)
\item 4 lines (one per environment): HotpotQA, APPS Intro, FEVER, WebShop
\item Shaded region for $N_{\text{explore}} \geq 30$ labeled
  ``stable zone''
\item Vertical dashed line at $N{=}50$ (our default)
\end{itemize}

\textbf{Panel (b): Gate threshold sensitivity.}
\begin{itemize}[nosep]
\item x-axis: $\tau \in \{0.3, 0.4, 0.5, 0.6, 0.7\}$
\item y-axis: SR (\%)
\item Same 4 environments as panel (a)
\item Vertical dashed line at $\tau{=}0.5$ (our default)
\end{itemize}

\textbf{Data source:} Run EAAG with each hyperparameter value,
hold other parameters at default. Record SR over 200 episodes
$\times$ 3 seeds.

\textbf{Visual design (ref: SEAG Figure~5 accuracy vs.\ cost curve):}
\begin{itemize}[nosep]
\item Line plot with markers, one color per environment
\item Error bars or shaded CI band (3-seed std)
\item Consistent y-axis range (both panels share y-axis)
\item Clean seaborn style
\end{itemize}

\textbf{Key message:} EAAG is robust to both $N_{\text{explore}}$
(stable for $\geq$30) and $\tau$ (stable across 0.3--0.7).}

EAAG is robust to $N_{\text{explore}} \in [30, 100]$ (stable SR for
$N_{\text{explore}} \geq 30$; marginal gains above 50) and gate
threshold $\tau \in [0.3, 0.7]$.

\subsection{Statistical Significance}

All head-to-head comparisons use 3-seed evaluation
(200 episodes/seed). Bootstrap confidence intervals (5000 resamples)
are reported for all comparisons. 18 of 34 wins are statistically
significant (95\% CI excludes zero). Pareto-dominance claims use a
conservative criterion that does not rely on p-values.

{\color{blue} \textbf{[DATA NEEDED: Table A8.]} Full statistical significance table.
For each method $\times$ environment pair: $\Delta$SR, 95\% bootstrap CI,
significant (Y/N). 48 cells total (6 baselines $\times$ 8 environments).}

\subsection{LLM Reasoning Prompts and Outputs}
\label{app:llm-prompts}

{\color{blue} \textbf{[DATA NEEDED.]} Full LLM prompt template used in
the Reason step of EAAG. Include one complete example showing:
(1) the prompt given to the LLM with exploration data summary,
(2) the LLM's full output including environment profile and
feature hypotheses, (3) which features LASSO subsequently selected.
Use WebShop as the example (most interesting LLM contribution).
This is important for reproducibility (NeurIPS checklist item).}


% ============================================================
% G. BROADER IMPACT
% ============================================================
\section{Broader Impact}
\label{app:impact}

This work addresses computational efficiency in LLM agent
deployment. By reducing unnecessary optimizer invocations, EAAG
lowers the energy and cost footprint of test-time compute without
sacrificing task performance. The prescriptive framework
(characterize information structure before designing gates) may
help practitioners avoid deploying fixed-direction methods that
actively harm performance in mismatched environments. We do not
foresee direct negative societal impacts beyond those inherent to
the LLM agents themselves.