% ============================================================
% 7. CONCLUSION
% ============================================================
\section{Conclusion}

<<<<<<< HEAD
The central claim of this paper is that \emph{uncertainty signals
are not sufficient statistics for the value of computation}.
Existing adaptive gating methods for LLM agents share an implicit
fixed-direction assumption---higher uncertainty indicates when
extra computation helps---which is the strict-monotone special
case of a structurally insufficient family: any policy depending
on the uncertainty signal alone, no matter how it is calibrated
or post-processed, must underperform never-gating on at least one
(environment, backbone) setting. The reason is that the value of
computation depends on a latent state property---information
poverty versus decision difficulty---that the signal alone cannot
identify and whose mixture is determined jointly by the environment
and the backbone. Across six environments and three model backbones,
signal--utility direction reverses along both axes, and once
misaligned, better calibration \emph{worsens} performance by more
precisely selecting harmful states. A two-source model grounded in
Simpson's paradox explains when and why reversal occurs, makes a
falsifiable prediction we verify on held-out data, and identifies
what an adaptive-compute method must condition on to recover
sufficiency: features that distinguish the two state types.
As an existence proof that this auxiliary conditioning---not gate
sophistication---is the bottleneck, DIAL augments the uncertainty
signal with a small pool of universal state features and fits a
sparse linear gate on signal-agnostic exploration data, adding
zero per-step inference overhead. Across every setting evaluated,
this minimal procedure already dominates fixed-direction
baselines, and its full range of gating behavior emerges from
direction learning alone.
=======
Existing adaptive gating methods for LLM agents assume that higher
uncertainty consistently indicates when extra computation helps.
We show this fixed-direction assumption fails: the signal--utility
relationship reverses across both environments and model backbones,
and when the assumed direction is wrong, better calibration worsens
performance by more precisely selecting harmful states. We trace
this reversal to a mixture of information-poor and
decision-difficult states whose composition varies across
(environment, backbone) pairs, and show that, under the two-source model, no policy depending on uncertainty alone can guarantee non-negative value across 
heterogeneous settings. DIAL learns the correct direction per setting by fitting
a sparse linear gate over a small set of state features, and
outperforms every fixed-direction baseline across six environments
and three backbones.
>>>>>>> 1bf5b0e2c05afb339eb3e76f610bbf4ff034c5a3
