% ============================================================
% 4. METHOD: EAAG
% ============================================================
\section{Method: Environment-Aware Adaptive Gating}
\label{sec:method}

<<<<<<< HEAD
Proposition~\ref{prop:necessity} establishes that direction
discovery is necessary for cross-environment robustness; no
existing method performs it (Table~\ref{tab:classification}).
EAAG (Environment-Aware Adaptive Gating) closes this gap by
learning both signal identity and direction from online
interaction data.
=======
Section~\ref{sec:signal-landscape} imposes three requirements on
any gating method that aims for cross-environment robustness.
(i)~Signal direction must be \emph{learned} per environment,
since Proposition~\ref{prop:necessity} rules out any fixed
choice. (ii)~The feature set must be \emph{multi-signal}, since
no single signal is reliably informative across environments
(\S\ref{sec:empirical-landscape}). (iii)~Exploration must be
\emph{signal-agnostic}, since conditioning on a signal whose
semantics are unknown would bias the collected data in a
direction that depends on the answer.
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf

EAAG is a deliberately minimal instantiation of these
requirements. Its contribution is not a new gate architecture
but a pipeline that reduces the adaptive-gating problem to a
form where a standard sparse linear model suffices: randomized
exploration (\S\ref{sec:explore}) handles the unknown prior
direction, an optional LLM feature layer (\S\ref{sec:reason})
proposes task-specific signals when universal ones are
inadequate, and a sparse linear gate (\S\ref{sec:learn})
recovers per-environment direction from the resulting dataset.
Figure~\ref{fig:eaag-pipeline} illustrates the full pipeline
and its connection to requirements~(i)--(iii).

<<<<<<< HEAD
\paragraph{Design rationale.}
The bottleneck in adaptive gating is direction, not model capacity
(\S\ref{sec:ablation}, Table~\ref{tab:capacity}): an MLP with the
wrong direction falls \emph{below} the no-trigger baseline, while
a logistic gate with the correct direction Pareto-dominates all
baselines. EAAG therefore pairs a rich discovery pipeline
(exploration + LLM reasoning + LASSO) with a deliberately simple
gate---a logistic function that encodes both signal identity and
direction.

{\color{red} \textbf{[FIGURE NEEDED: Figure 4.]} EAAG method overview diagram.
Three-phase pipeline: Explore (randomized gating, collect D) $\to$
Reason (LLM analyzes D, generates environment profile + feature hypotheses)
$\to$ Learn (LASSO selects features, trains logistic gate with signed weights).
Show arrows from \S3 findings (direction reversal, signal identity, necessity)
to each design choice.
Similar to FLARE's method section which has equations inline but
a clear conceptual flow. Consider a horizontal pipeline diagram.}
=======
\paragraph{Why a minimal gate suffices.}
The real difficulty identified in Section~\ref{sec:signal-landscape}
is identifying signal direction per environment, not
classification capacity. Once exploration and feature selection
have produced unbiased $(s, U)$ data and a usable feature pool,
the remaining task is well-posed for a standard sparse linear
model. On HotpotQA (Table~\ref{tab:capacity}), a logistic gate,
an MLP, and a $2560$-dimensional hidden-state probe all reach
$\sim$$95\%$ SR \emph{with the correct direction}; under a
reversed direction, the MLP falls \emph{below} the no-trigger
baseline because a more expressive gate more precisely targets
harmful states. In the environments we study, the marginal
return on gate complexity is small compared to the return on
getting direction right, so we invest complexity in exploration
and feature selection and keep the deployed gate minimal.

<<<<<<< HEAD
{\color{red} \textbf{[FIGURE NEEDED: Figure~\ref{fig:eaag-pipeline}.]}
EAAG pipeline diagram. \textbf{Layout:} Horizontal three-stage pipeline
with requirement mapping above and signed-weight example below.

\textbf{Top row (requirements):} Three boxes labeled
``Req~(iii): Signal-agnostic collection'',
``Req~(ii): Multi-signal features'',
``Req~(i): Direction must be learned'',
each with a downward arrow to the corresponding pipeline stage.

\textbf{Middle row (pipeline):} Three colored boxes connected by arrows:
\textsc{Explore} (blue) $\to$ \textsc{Reason} (orange, dashed border for optional)
$\to$ \textsc{Learn} (green) $\to$ \textsc{Deploy} (gray).
Inside each box, 1--2 line summary:
Explore: ``$\varepsilon{=}0.5$ random gating, 50 eps $\to$ $\mathcal{D}$'';
Reason: ``LLM proposes task features $\to$ $\phi_{\text{cand}}$'';
Learn: ``LASSO $\to$ gate $g(s)$ with signed $\mathbf{w}$'';
Deploy: ``Zero overhead per step''.

\textbf{Bottom annotation:} Two small boxes showing example signed weights:
HotpotQA ($w_{\text{entropy}}{=}{-}0.34$, Type~I: high entropy $\to$ \emph{don't} trigger)
vs.\ APPS Interview ($w_{\text{entropy}}{=}{+}0.32$, Type~D: high entropy $\to$ trigger).
Use red/blue coloring to indicate direction.

\textbf{Style:} Clean TikZ boxes similar to CATTS Figure~1 and CoRefine Figure~3.
Rounded rectangles, minimal text, consistent color scheme.
See drawing prompt in \texttt{figure\_prompts.md}.}

=======
{\color{red} \textbf{[FIGURE NEEDED: Figure 4.]} EAAG pipeline
diagram: Explore $\to$ Reason (optional) $\to$ Learn, with
arrows from requirements (i), (ii), (iii) to the corresponding
stages.}
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf
>>>>>>> a7e30fbd74c674828774fd4253d874b033355a26

\subsection{Explore: Signal-Agnostic Data Collection}
\label{sec:explore}

Exploration collects the $(s, U)$ pairs that will be used to fit
the gate. The central constraint is requirement~(iii): the
collection policy must not condition on any signal whose
semantics we are trying to learn. Any signal-dependent
alternative, such as uncertainty sampling, would condition on a
signal whose relationship to $U$ is exactly what
\S\ref{sec:empirical-landscape} showed to be unknown, biasing
$\mathcal{D}$ in a direction that depends on the quantity being
estimated. Warm-starting from prior-method data inherits the
same circularity.

EAAG therefore uses the simplest policy consistent with this
constraint. At each step~$t$ in episode~$i$, a Bernoulli coin
flip determines the gating decision:
\begin{equation}
  z_t^{(i)} \sim \mathrm{Bernoulli}(\varepsilon_{\text{explore}}),
  \quad \varepsilon_{\text{explore}} = 0.5,
  \label{eq:explore}
\end{equation}
where $z_t{=}1$ triggers the optimizer~$T$ and $z_t{=}0$ uses
the base action. Over $N_{\text{explore}} = 50$ episodes, this
yields an unbiased dataset
\begin{equation}
  \mathcal{D} = \bigl\{
    \bigl(\phi(s_t^{(i)}),\; U_t^{(i)}\bigr)
    : z_t^{(i)} = 1
  \bigr\}_{t, i},
  \label{eq:dataset}
\end{equation}
where $U_t^{(i)} \in \{0, 1\}$ indicates whether the optimizer
improved the outcome at step~$t$, and $\phi(s)$ is the feature
vector described in \S\ref{sec:reason}.


\subsection{Reason: LLM Feature Layer}
\label{sec:reason}

The next step builds the feature pool $\phi_{\text{cand}}(s)$
that the gate will be fit over. Requirement~(ii) demands that
this pool be rich enough to include each environment's relevant
signals, but not so environment-specific that deployment
requires manual engineering per task.

EAAG's default is a fixed set of universal features:
\begin{equation}
  \phi_{\text{univ}}(s) = [\,\texttt{step\_count},\;
  \texttt{token\_entropy},\; \texttt{state\_length},\;
  \texttt{action\_entropy},\; \ldots\,],
  \label{eq:universal}
\end{equation}
available in every environment without task-specific engineering.
In the environments we study, this pool alone is sufficient to
recover a gate that outperforms all fixed-direction baselines
(\S\ref{sec:ablation}).

For environments where universal features may be inadequate,
such as WebShop, whose gating benefits from task-specific
signals like \texttt{num\_available\_actions}, EAAG offers an
\emph{optional} automation layer: an LLM is given a summary of
$\mathcal{D}$ and proposes task-specific feature hypotheses
$\phi_{\text{LLM}}(s)$, which join the universal pool to form
\begin{equation}
  \phi_{\text{cand}}(s)
  = \phi_{\text{univ}}(s) \;\cup\; \phi_{\text{LLM}}(s).
  \label{eq:cand}
\end{equation}
The ablation in \S\ref{sec:ablation} shows its gains concentrate
in the few settings where universal features miss the relevant
structure, so the LLM is best viewed as an automation
convenience for zero-shot deployment rather than a core
component the method depends on.


\subsection{Learn: Sparse Linear Gate}
\label{sec:learn}

Given $\mathcal{D}$ and $\phi_{\text{cand}}$, EAAG fits an
$\ell_1$-regularized logistic regression:
\begin{equation}
  \mathbf{w}^*, b^* = \argmin_{\mathbf{w}, b}
  \sum_{(\phi, U) \in \mathcal{D}}
  \ell\!\bigl(\sigma(\mathbf{w}^\top \phi + b),\, U\bigr)
  + \lambda \|\mathbf{w}\|_1,
  \label{eq:lasso}
\end{equation}
where $\ell$ is the binary cross-entropy loss and
$\lambda$ controls sparsity over $\phi_{\text{cand}}$ so that
uninformative features drop out. The deployed gate is
\begin{equation}
  g(s) = \mathbf{1}\!\left[\sigma(\mathbf{w}^{*\top} \phi(s) + b^*)
  > \tau\right],
  \label{eq:gate}
\end{equation}
with threshold $\tau$ set by cross-validation on $\mathcal{D}$.
Gate fitting completes in under a second, and deployment adds
zero per-step overhead beyond feature computation.
Implementation details are in
Appendix~\ref{app:eaag-details}.

\paragraph{Signed weights as a diagnostic.}
The $\ell_1$ penalty in Eq.~\ref{eq:lasso} simultaneously
performs feature selection and direction discovery. The sign of
each fitted weight provides an interpretable per-environment
estimate of the Two-Source mixture of \S\ref{sec:toy-model}:
$w_i > 0$ indicates the environment sits on the Type~D side of
the reversal threshold $p_I^*$ for signal~$i$ (high signal
$\to$ trigger), $w_i < 0$ the Type~I side (high signal $\to$
\emph{don't} trigger), and $w_i \approx 0$ an uninformative
signal. This is a diagnostic reading of the fitted model, not a
novel modeling choice; any signed linear classifier would expose
the same information.

\paragraph{Online adaptation.}
For drifting environments, EAAG can optionally refit during
deployment using $\varepsilon$-greedy exploration
($\varepsilon: 0.1 \to 0$) and periodic retraining. This mode
is \emph{not free}: $\varepsilon$-greedy overrides ignore the
current gate, so in Type~I environments a fraction of
overridden steps will trigger harmful rollouts. For stationary
environments, which are the setting of our main experiments,
we recommend skipping online adaptation and relying on the
offline-fitted gate.

% --- Algorithm Box ---
\begin{algorithm}[t]
\caption{EAAG: Environment-Aware Adaptive Gating}
\label{alg:eaag}
\small
\begin{algorithmic}[1]
\Require Environment~$E$, agent~$M$, optimizer~$T$,
  $N_{\text{explore}}$ episodes, $\varepsilon_{\text{explore}}{=}0.5$
\Ensure Fitted gate $g(s)$
\Statex
\Statex \textbf{Phase 1: Explore} \hfill\textit{// Req~(iii)}
\State $\mathcal{D} \gets \emptyset$
\For{episode $i = 1, \ldots, N_{\text{explore}}$}
  \For{step $t$ in episode $i$}
    \State $z_t \sim \mathrm{Bernoulli}(\varepsilon_{\text{explore}})$
    \State Execute $T$ if $z_t{=}1$, else base action
    \State $\mathcal{D} \gets \mathcal{D} \cup
      \{(\phi(s_t),\, U_t,\, z_t)\}$
  \EndFor
\EndFor
\Statex
\Statex \textbf{Phase 2: Reason} (optional) \hfill\textit{// Req~(ii)}
\State $\phi_{\text{univ}} \gets
  [\texttt{step\_count}, \texttt{entropy}, \ldots]$
\State $\phi_{\text{LLM}} \gets
  \textsc{LLM\_Propose}(\mathrm{summary}(\mathcal{D}))$
\State $\phi_{\text{cand}} \gets
  \phi_{\text{univ}} \cup \phi_{\text{LLM}}$
\Statex
\Statex \textbf{Phase 3: Learn} \hfill\textit{// Req~(i)}
\State $\mathbf{w}^*, b^* \gets
  \argmin_{\mathbf{w},b}
  \sum \ell(\sigma(\mathbf{w}^\top\phi + b), U)
  + \lambda\|\mathbf{w}\|_1$
\State $g(s) \gets
  \mathbf{1}[\sigma(\mathbf{w}^{*\top}\phi(s) + b^*) > \tau]$
\Statex
\State \Return $g(s)$
  \hfill\textit{// Deploy: zero per-step overhead}
\end{algorithmic}
\end{algorithm}
