\section{Related Work}
\label{sec:related}
Prior work on adaptive test-time compute uses diverse mechanisms, but shares a common assumption: gating signals have a fixed, setting-independent relationship with compute value. We organize the literature by how each line embeds this assumption.

\paragraph{Gating on uncertainty signals.}
Signal-based methods allocate extra computation when token confidence, entropy, or disagreement exceeds a threshold, encoding the fixed-direction assumption directly in their trigger rule. In reasoning settings, this takes the form of extended chain-of-thought generation~\cite{seag2025, cats2025, corefine2026, thinkjustenough2025, s1_2025, han2025tokenbudget}. In agent settings, vote-based methods~\cite{catts2026} invoke an arbiter when candidate actions disagree, and verbalized uncertainty methods~\cite{auq2026} trigger reflection when self-reported confidence is low. All of these methods hardcode the direction in which uncertainty maps to compute need; we show this direction is itself a function of (environment, backbone) and cannot be hardcoded.

\paragraph{Gating on learned difficulty.}
Hidden-state probes~\cite{diffadapt2025, zhu2025llmalreadyknows} estimate task difficulty from internal representations, and RL-based methods~\cite{adaptthink2025, thinkless2025, aggarwal2025l1, paglieri2025learning} learn think/no-think policies end-to-end. Although these methods commit to neither a specific signal nor an explicit threshold, their evaluations primarily focus on homogeneous reasoning benchmarks such as GSM8K~\cite{DBLP:journals/corr/abs-2110-14168} and GPQA~\cite{DBLP:journals/corr/abs-2311-12022}, where higher difficulty is typically aligned with greater benefit from additional reasoning. The difficulty-compute relationship is thus never tested for reversal. We test it explicitly in heterogeneous agent environments, where it fails.

\paragraph{Confidence calibration vs.\ utility calibration.}
Classical confidence calibration aligns a model's stated confidence with its action correctness; recent work shows that calibration quality is domain-dependent~\cite{tao2025revisiting, heo2025llmuncertainty, yadkori2024believe}. Our problem is distinct: \emph{utility calibration} aligns gating signals with the value of additional computation, not action correctness. Rational metareasoning~\cite{russell1991right, desabbata2024metareasoning} formalizes this VOC target but classically assumes $\text{VOC} \geq 0$; we show this fails when the same model both proposes and evaluates actions (\S\ref{sec:toy-model}). Beyond domain-dependent quality, the \emph{sign} of the signal-to-utility correlation reverses across (environment, backbone) pairs, so even correctness-calibrated confidence is insufficient for gating. Full discussion in Appendix~\ref{app:related}.