% ============================================================
% ABSTRACT
% ============================================================
\begin{abstract}
Large language model (LLM) agents benefit substantially from
test-time compute, but applying expensive optimization to every
decision step is wasteful. Exist gating methods address this by using uncertainty
signals to decide when extra compute is worthwhile. 
However, these methods share an implicit assumption: a fixed mapping from signal value to compute benefit. 
We show that this assumption is wrong. Across eight diverse
agent environments and three model backbones, we find that the
signal--utility direction \emph{reverses}: the same signal value
that predicts rollout benefit in one environment predicts rollout
harm in another. We trace this to a conflation of two uncertainty
sources, \emph{information poverty} and \emph{decision difficulty},
and formalize it via a two-source model grounded in Simpson's
paradox, where subgroup-level trends reverse upon aggregation. 
We prove that direction discovery is a necessary condition for non-negative value of computation across environments. Motivated by this result, we propose EAAG (Environment-Aware Adaptive Gating), which combines exploration, LLM-based reasoning, and sparse direction learning to discover environment-specific gating patterns. Across eight environments, EAAG outperforms all fixed-direction baselines with zero per-step overhead, and its adaptive gating behavior emerges from direction learning alone.
\end{abstract}
