% ============================================================
% 7. CONCLUSION
% ============================================================
\section{Conclusion}
All existing adaptive gating methods for LLM agents rest on an
implicit assumption: that uncertainty signals have a fixed
relationship with compute value. We show that this assumption
<<<<<<< HEAD
reflects a structural limitation---uncertainty signals are not
=======
reflects a structural limitation: uncertainty signals are not
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf
sufficient statistics for the value of computation, because their
meaning depends on a latent state property (information poverty
vs.\ decision difficulty) that the signal alone cannot identify.
Across eight environments and three model backbones, signal--utility
direction reverses; once misaligned, improved calibration
\emph{worsens} performance by more precisely selecting harmful
states. A two-source model grounded in Simpson's paradox explains
when and why reversal occurs, and a necessity proof establishes
that direction discovery is unavoidable for cross-environment
<<<<<<< HEAD
robustness. EAAG instantiates this principle through exploration,
LLM-based reasoning, and sparse direction learning, with zero
additional per-step overhead. Across all evaluated settings,
EAAG dominates fixed-direction baselines, with the full range of
gating behavior---from aggressive triggering to near-total
suppression---emerging from direction learning alone.
=======
robustness. EAAG instantiates this principle with a minimal
pipeline that collects unbiased data through randomized
exploration and recovers per-environment direction through sparse
linear learning, adding zero per-step overhead at deployment.
Across all evaluated settings, EAAG dominates fixed-direction
baselines, with the full range of gating behavior, from aggressive
triggering to near-total suppression, emerging from direction
learning alone.
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf
