% ============================================================
% 5. EXPERIMENTS
% ============================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Experimental Setup}
\label{sec:setup}

We evaluate EAAG along three axes: does environment-aware gating
Pareto-dominate fixed-direction baselines
(\S\ref{sec:main-results}); which components drive the gains
(\S\ref{sec:ablation}); and do the Two-Source Model's predictions
hold on held-out data (\S\ref{sec:theory-verification}).

\paragraph{Environments.} We evaluate on 8 environments spanning QA,
code generation, web navigation, fact verification, text games, and
crafting planning (Table~\ref{tab:env-setup}): HotpotQA,
APPS Intro, WebShop, FEVER, TWExpress, Plancraft, APPS Interview,
and CRUXEval. These environments are selected to cover the full
range of the Two-Source Model, from pure Type~I (FEVER) through
mixed ($\approx p_I^*$, APPS Intro) to Type~D (APPS Interview),
including extreme rollout properties (TWExpress: rollout-safe;
Plancraft: rollout-harmful).

\paragraph{Baselines.} All methods share the same agent and
optimizer $T$; only the gating decision differs. Bounds:
base\_only (never trigger), always\_trigger, oracle.
Fixed-direction baselines: CaTS, SEAG, CoRefine, CATTS, AUQ,
<<<<<<< HEAD
s1\_budget---all assuming high uncertainty warrants compute.
=======
s1\_budget, all assuming high uncertainty warrants compute.
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf
Ablation: BSW (direction intentionally reversed). Full
specifications in Appendix~\ref{app:experiments}.

\paragraph{Metrics.} SR (success rate, $\uparrow$) and Cost
(rollouts per episode during deployment, $\downarrow$).
A method Pareto-dominates another if $\text{SR} \geq$ and
$\text{Cost} \leq$ with at least one strict inequality.
Cost measures only the \emph{deployment} rollout budget; calibration
or exploration overhead is excluded for all methods, ensuring a fair
comparison of runtime efficiency.

\begin{table}[t]
\caption{Environment setup. Base/Always: SR without/with optimizer.
$\Delta$: rollout headroom.}
\label{tab:env-setup}
\centering\small
\begin{tabular}{lcccl}
\toprule
Environment & Base SR & Always SR & $\Delta$ & Optimizer $T$ \\
\midrule
HotpotQA      & 49.0\% & 97.0\% & +48.0 & Per-action eval \\
APPS Intro    & 58.5\% & 64.5\% &  +6.0 & $K$-variant sampling \\
WebShop       &  7.2\% & 43.0\% & +35.8 & LLM-Propose-$K$ \\
FEVER         & 37.0\% & 99.8\% & +62.8 & Per-action eval \\
TWExpress     & 67.5\% & 99.3\% & +31.8 & Per-action eval \\
Plancraft     & 29.8\% & 22.8\% &  $-$7.0 & Per-action eval \\
APPS Interview& 60.5\% & 79.5\% & +19.0 & $K$-variant sampling \\
CRUXEval      & 85.0\% & 99.5\% & +14.5 & $K$-variant sampling \\
\bottomrule
\end{tabular}
\end{table}


\subsection{Main Results}
\label{sec:main-results}

\textbf{EAAG Pareto-dominates fixed-direction baselines.}
Table~\ref{tab:main} reports SR and cost across 8 environments
(summary in Table~\ref{tab:winloss}). EAAG matches or exceeds
every fixed-direction baseline in SR while using fewer rollouts;
the two exceptions occur in ceiling-effect environments where EAAG
already approaches the oracle. No calibration data is required
(statistical details in Appendix~\ref{app:additional}). Beyond
aggregate numbers, EAAG exhibits \emph{qualitatively different}
gating behavior: its trigger rate, direction, and signal weights
all adapt to the environment, as we analyze below.

{\color{blue} \textbf{[DATA NEEDED: Table~\ref{tab:main}.]} Main results table.
Full method $\times$ environment grid. Rows = methods (base\_only,
always\_trigger, oracle, CaTS, SEAG, CoRefine, CATTS, AUQ, s1\_budget,
BSW, EAAG). Columns = 8 environments, each with SR(\%) and Cost (ro/ep).
Bold best SR per environment. $\dagger$ marks calibration-dependent methods.

\textbf{Format reference:} SEAG Table~1 (benchmark $\times$ method $\times$ backbone grid)
and CoRefine Table~1 (mean$_{\pm\text{std}}$ over 5 runs).
Include $\pm$ std if available. Add $\Delta$ column showing SR improvement over base.}

{\color{blue} \textbf{[DATA NEEDED: Table~\ref{tab:winloss}.]} Win/loss summary
over all baseline-environment pairs. Rows = baselines, columns = number
of environments where EAAG wins/ties/loses on SR and on Cost.}

{\color{red} \textbf{[FIGURE NEEDED: Figure~\ref{fig:pareto}.]}
\textbf{Pareto frontier plot} (ref: CATTS Figure~5).

\textbf{Layout:} $2 \times 3$ grid of subplots (6 representative environments:
HotpotQA, APPS Intro, WebShop, FEVER, Plancraft, APPS Interview).
x-axis = Cost (rollouts/episode), y-axis = SR(\%).

\textbf{Data:} From \texttt{tab\_main\_results/} CSV files.
Each method is one point: (cost, SR).

\textbf{Visual design (ref: CATTS Figure 5):}
\begin{itemize}[nosep]
\item Different marker shapes by method category:
  $\circ$ = bounds (base, always, oracle),
  $\triangle$ = fixed-direction baselines (CaTS, SEAG, CoRefine, CATTS, AUQ, s1),
  $\bigstar$ = EAAG (larger, prominent)
\item Color palette: bounds in gray, fixed-direction in a sequential colormap
  (e.g., viridis or tab10), EAAG in red/crimson
\item Draw Pareto frontier as dashed line connecting non-dominated points
\item Mark EAAG with a star ($\bigstar$) like CATTS does
\item Add light shading below Pareto frontier
\item Shared legend across subplots
\end{itemize}

\textbf{Key message:} EAAG sits on or near the Pareto frontier in every
environment; fixed-direction baselines cluster below in Type~I environments.

\textbf{Improvement over current fig2\_pareto.png:}
Current version is too cluttered, markers too small, no Pareto line.
Add: (1) Pareto frontier dashed line, (2) larger star marker for EAAG,
(3) method-category marker shapes, (4) consistent axis ranges per panel.}

\paragraph{Cross-environment patterns.}
The results align with the Two-Source Model along three axes.
The gap between EAAG and baselines is largest where direction
mismatch is most severe: in Type~I environments, all ``high
uncertainty $\to$ trigger'' baselines fall below the no-trigger
<<<<<<< HEAD
baseline because the assumed direction is wrong---no amount of
=======
baseline because the assumed direction is wrong, and no amount of
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf
calibration rescues them. Conversely, EAAG's advantage is smallest
in high-headroom environments where most methods approach the
oracle regardless of direction. Most tellingly, EAAG adapts not
only direction but \emph{gating magnitude}: aggressive triggering
in rollout-safe environments (TWExpress, RR\,=\,73\%),
moderate triggering in mixed environments, and step-dependent
<<<<<<< HEAD
decay in rollout-harmful ones (Plancraft: 49\% at step~0 $\to$
$<$20\% later). This full behavioral range---from near-always
to near-never triggering---emerges from direction learning
alone, without explicit headroom estimation.
=======
decay in rollout-harmful ones (Plancraft: 49\% at step~0 to
$<$20\% later). This full behavioral range, from near-always
to near-never triggering, emerges from direction learning alone,
without explicit headroom estimation.
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf

{\color{red} \textbf{[FIGURE NEEDED: Figure~\ref{fig:entropy-bin}.]}
\textbf{SR by entropy bin} (ref: SEAG Figure~4 + CATTS Figure~4).

\textbf{Layout:} $1 \times 3$ panels, each showing one representative
environment type: (a) Type~I (FEVER or HotpotQA),
(b) Mixed (APPS Intro), (c) Type~D (APPS Interview).

\textbf{Data:} From exploration data $\mathcal{D}$.
For each environment:
\begin{enumerate}[nosep]
\item Bin steps by entropy into 4--5 bins (e.g., 0--0.3, 0.3--0.6, 0.6--0.9, 0.9+)
\item For each bin, compute SR of: EAAG, best fixed-direction baseline,
  always\_trigger, base\_only
\item Also compute: proportion of steps in each bin (annotate on top like SEAG Fig~4)
\end{enumerate}

\textbf{Visual design (ref: SEAG Figure 4):}
\begin{itemize}[nosep]
\item Grouped bar chart: 3--4 methods per entropy bin
\item Annotate proportion of steps at top of each bin group (like SEAG)
\item Color: EAAG = red/crimson, baselines = blue/green/gray
\item Show numbers on top of each bar
\end{itemize}

\textbf{Key message:}
\begin{itemize}[nosep]
\item In Type~I panel: At high entropy, EAAG maintains SR while baselines
  degrade (they trigger when they shouldn't)
\item In Type~D panel: At high entropy, both EAAG and baselines improve
  (direction aligns), but EAAG is more selective
\item In Mixed panel: EAAG adapts within-episode
\end{itemize}

\textbf{Data source:} Compute from per-step exploration logs.
Fields needed: \texttt{env, step, entropy, trigger\_decision, U, method}.}


\subsection{Ablation Studies}
\label{sec:ablation}

The main results show \emph{that} EAAG works; we now ask
\emph{why}. We ablate the two components of EAAG that \S\ref{sec:signal-landscape}
identified as necessary: direction learning and the LLM feature
layer.

\textbf{Direction is the primary determinant of gate quality.}
Intentionally reversing the learned direction (BSW ablation) causes
catastrophic failure across environments, with an MLP gate using the
wrong direction falling \emph{below} the no-trigger baseline. This
confirms the central prediction of Proposition~\ref{prop:necessity}:
direction is not a minor calibration detail but the primary
determinant of gate quality. The magnitude of degradation correlates
<<<<<<< HEAD
with signal strength (Appendix~\ref{app:wrong-direction}), as the Two-Source
Model predicts.
=======
with signal strength (Appendix~\ref{app:wrong-direction}), as the
Two-Source Model predicts.
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf

\textbf{The LLM feature layer is an automation convenience, not
a core component.} Removing the LLM layer changes SR by
${<}$1\,pp in most environments; the gains concentrate in
settings like WebShop where task-specific signals meaningfully
extend the universal feature pool. In one environment LLM
features even introduce noise. Consistent with the framing in
\S\ref{sec:reason}, the LLM is best viewed as an automation
convenience that enables zero-shot deployment on new environments
rather than a component the method's main claims depend on.

{\color{red} \textbf{[FIGURE NEEDED: Figure~\ref{fig:trigger-adapt}.]}
\textbf{Environment-adaptive trigger behavior} (ref: CATTS Figure~2 trajectory profiles).

\textbf{Layout:} $2 \times 3$ grid, 6 environments.
Each subplot: x-axis = step index, y-axis = EAAG trigger rate (\%).

\textbf{Data:} From deployment-phase gate decisions.
For each environment and step index $t$:
\begin{enumerate}[nosep]
\item Compute trigger rate = \% of episodes where $g(s_t) = 1$
\item Compute 95\% CI band (bootstrap or binomial)
\item Annotate overall trigger rate (RR) in corner
\end{enumerate}

\textbf{Visual design (ref: CATTS Figure~2 dual-color line):}
\begin{itemize}[nosep]
\item Solid blue line with shaded 95\% CI band
\item Horizontal dashed gray line at 50\% for reference
\item Annotate overall RR in each panel corner
\item Order panels by $\Delta$ (rollout headroom): TWExpress first, Plancraft last
\end{itemize}

\textbf{Key message:} Trigger rate varies from 73\% (TWExpress, rollout-safe)
to $<$20\% late-step (Plancraft, rollout-harmful). Step-dependent decay
in rollout-harmful environments. This behavioral range emerges from
direction learning alone.

\textbf{Data source:} \texttt{fig\_trigger\_rate/} CSV files.
Current fig\_trigger\_rate\_by\_step.png exists but needs: (1) consistent
axis range, (2) panel ordering by $\Delta$, (3) cleaner style matching
other figures.}

\textbf{Gating magnitude emerges from direction learning alone.}
The gate's trigger rate varies from 73\% (TWExpress) to $<$20\%
(late-step Plancraft), with step-dependent decay in
rollout-harmful environments. These patterns are not explicitly
<<<<<<< HEAD
programmed: they arise from logistic regression over
LASSO-selected features, whose signed weights implicitly encode
each environment's signal--utility structure.
=======
programmed: they follow from the per-environment signed weights
fit in \S\ref{sec:learn}, which capture each environment's
signal--utility structure.
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf

{\color{red} \textbf{[FIGURE NEEDED: Figure~\ref{fig:direction-cost}.]}
\textbf{Cost of wrong direction scales with signal strength}
(ref: CATTS Figure~3 bar chart + Figure~4 net advantage).

\textbf{Layout:} Single panel scatter plot.
x-axis = signal strength $|\rho|$ per environment,
y-axis = BSW degradation (pp SR drop vs.\ EAAG with correct direction).

\textbf{Data:} From \texttt{fig3\_bsw\_direction/} and
\texttt{tab\_signal\_discovery/} CSV files.
For each environment:
\begin{enumerate}[nosep]
\item $|\rho|$ = absolute Spearman correlation of best signal
  (from signal discovery table)
\item BSW\_drop = SR(EAAG) $-$ SR(BSW) in percentage points
\end{enumerate}

\textbf{Visual design:}
\begin{itemize}[nosep]
\item Scatter with environment labels next to each point
\item Color by environment type: blue = Type~I, red = Type~D,
  gray = mixed/weak
\item Dashed trend line (linear fit) with $r$ value
\item Large, clear font for environment labels
\end{itemize}

\textbf{Key message:} Wrong direction is more damaging when signal is
stronger. Environments with $|\rho| > 0.3$ see $>$15pp degradation.
Directly validates Proposition~1: direction is the bottleneck.

\textbf{Improvement over current fig3\_bsw\_cost.png:}
Current version works but needs: (1) Type I/D color coding,
(2) larger text, (3) clearer trend line annotation.}


\subsection{Two-Source Model Verification}
\label{sec:theory-verification}

The ablations confirm EAAG's practical advantage; we now test
whether the Two-Source Model (\S\ref{sec:toy-model}) makes correct
predictions on held-out data. Following
\citet{schaeffer2023emergent}, we derive three testable predictions
and verify each:
\begin{enumerate}[nosep,label=\textbf{P\arabic*:}]
\item \textbf{Temporal dynamics.} Within the same environment,
  $\rho$ should \emph{decrease} over the episode: early steps mix
  Type~I and Type~D uncertainty, while late steps, after
  information-gathering attempts, isolate the residual Type~I
  component.
\item \textbf{Cross-environment consistency.} Environments with
  similar task structure should exhibit similar $\rho$
  (e.g., FEVER $\approx$ HotpotQA, both search-based QA).
\item \textbf{Signal identity alignment.} In Type~I environments,
  the strongest signal should measure information sufficiency; in
  Type~D environments, decision complexity.
\end{enumerate}

{\color{red} \textbf{[FIGURE NEEDED: Figure~\ref{fig:temporal-shift}.]}
\textbf{P1 verification: temporal dynamics of $\rho$}
(ref: CATTS Figure~2 dual-panel trajectory + current fig\_p1\_temporal\_shift.png).

\textbf{Layout:} Single panel, grouped bar chart.
x-axis = environments (HotpotQA, FEVER, TWExpress, APPS, WebShop).
Two bars per environment: early steps (blue) vs.\ late steps (red).
y-axis = $\rho$(entropy, utility).

\textbf{Data:} From \texttt{fig\_p1\_temporal\_shift/} CSV files.
\begin{enumerate}[nosep]
\item Split each environment's trajectory data at median step
\item Compute $\rho$ for early and late subsets
\item Compute 95\% CI via bootstrap
\end{enumerate}

\textbf{Visual design:}
\begin{itemize}[nosep]
\item Paired bar chart with error bars
\item Annotate $\Delta = \rho_{\text{late}} - \rho_{\text{early}}$
  above each pair
\item Horizontal dashed line at $\rho = 0$
\end{itemize}

\textbf{Key message:} $\rho$ consistently shifts more negative
from early to late steps, confirming that late steps isolate
the dominant uncertainty type.

\textbf{Current fig\_p1\_temporal\_shift.png is good;} minor
improvements: (1) add $\Delta$ annotations, (2) consistent error bars.}

\paragraph{P1: Temporal dynamics (confirmed).}
We split trajectory data into early (step $\leq$ median) and late
(step $>$ median) subsets. In HotpotQA, $\rho$ shifts from $-0.176$
(early) to $-0.437$ (late): late-step entropy reflects deeper
information poverty after failed evidence retrieval. In APPS Intro,
$\rho$ shifts from $+0.102$ (early, non-significant) to $-0.144$
(late, $p{=}0.04$): early exploration diversity gives way to
debugging frustration. In WebShop, the strong positive early-phase
signal ($\rho{=}{+}0.285$) vanishes in the late phase. The pattern
is consistent across all environments with sufficient signal (full
<<<<<<< HEAD
data in Appendix~\ref{app:model-derivation}): early steps mix both uncertainty
sources, while late steps isolate the dominant type.
=======
data in Appendix~\ref{app:model-derivation}): early steps mix both
uncertainty sources, while late steps isolate the dominant type.
>>>>>>> cc1bc6e4f1be11f03658207b43893225ca507faf

\paragraph{P2: Cross-environment consistency (confirmed).}
We compare same-family pairs. FEVER ($\rho{=}{-}0.119$) and HotpotQA
($\rho{=}{-}0.041$), both search-based QA tasks requiring evidence
retrieval, share negative entropy--utility direction.
APPS Intro ($\rho{=}{+}0.012$) and APPS Interview
($\rho{=}{+}0.317$), both code generation tasks, share non-negative
direction. The magnitude difference within each pair reflects the
degree of Type~I contamination; FEVER has more early-step
information poverty than HotpotQA.

\paragraph{P3: Signal identity alignment (confirmed).}
We examine the sparse-selected features per environment.
In Type~I environments (HotpotQA, FEVER), the strongest signal
(\texttt{step\_count}, $|\rho|{=}0.494$ and $0.619$) measures
trajectory progress, a proxy for information accumulation.
In WebShop (mixed, choice-dominated),
\texttt{num\_available\_actions} ($\rho{=}{+}0.444$) measures the
decision space size, a proxy for decision complexity. All three
predictions are confirmed, providing converging evidence for the
Two-Source Model.

As a supplementary test, we manipulate information availability
within HotpotQA itself (InfoPoor vs.\ InfoRich conditions) and
confirm that the identity of the most informative signal shifts
accordingly, consistent with the signal-identity finding in
\S\ref{sec:empirical-landscape}. Full results are reported in
Appendix~\ref{app:controlled}.


\subsection{Robustness of Direction Reversal}
\label{sec:robustness}

Is direction reversal a genuine phenomenon or a confounding
artifact? We rule this out through three analyses, plus
multi-backbone verification (Appendix~\ref{app:multi-backbone})
showing that direction depends on the
\emph{environment~$\times$~model} interaction, not on either
factor alone.

\paragraph{Stratified analysis: controlling for trajectory length.}
Trajectory length varies across environments and correlates with both
entropy and utility. To rule out length as a confounder, we stratify
by step count and recompute $\rho(\text{entropy}, U)$ within each
stratum.
\emph{Result}: Direction reversal persists within trajectory-length
strata where utility variance is non-zero. In HotpotQA, $\rho$ remains
negative across all three strata ($-0.18$/$-0.46$/$-0.42$).
The reversal is not an artifact of trajectory length.

\paragraph{Interventional evidence.}
The BSW ablation (\S\ref{sec:ablation}) provides
\emph{interventional} evidence: reversing the gate's learned
direction causes catastrophic SR degradation across environments,
with the MLP gate falling below the no-trigger baseline. If
direction reversal were merely correlational, this manipulation
should have no systematic effect. The observed damage confirms
that direction encodes a causal relationship.

\paragraph{Gate complexity ablation: direction outweighs capacity.}
To verify that the bottleneck is direction rather than gate
complexity, we compare gate architectures with correct vs.\ wrong
direction (Table~\ref{tab:capacity}).

\begin{table}[t]
\caption{Gate complexity ablation on HotpotQA. Direction matters
more than model capacity: a logistic gate with correct direction
outperforms an MLP with wrong direction by 49.9\,pp.}
\label{tab:capacity}
\centering\small
\begin{tabular}{lccc}
\toprule
Gate & Direction & SR (\%) & $\Delta$ \\
\midrule
Logistic (5 feat) & Correct & 95.2 & +46.2 \\
MLP (5 feat)      & Correct & $\sim$95 & $\sim$+46 \\
Hidden state (2560-d) & Correct & $\sim$95 & $\sim$+46 \\
\midrule
Logistic (5 feat) & Wrong & 62.0 & +13.0 \\
MLP (5 feat)      & Wrong & 45.3 & $-$3.7 \\
\midrule
No gate (base)    & n/a & 49.0 & 0 \\
\bottomrule
\end{tabular}
\end{table}

With correct direction, all architectures achieve $\sim$95\% SR;
the jump from logistic to MLP to hidden-state probe adds $<$1\,pp.
With wrong direction, higher capacity makes performance \emph{worse}
(MLP 45.3\% $<$ logistic 62.0\%), because the more powerful gate
more precisely targets harmful states. Across the gate architectures
we tested, signal direction consistently outweighs both signal count
and gate complexity as a determinant of gate quality.

{\color{red} \textbf{[FIGURE NEEDED: Figure~\ref{fig:case-study}.]}
\textbf{Trajectory-level gate decision visualization}
(ref: CoRefine Figure~4 tree decision visualization).

\textbf{Layout:} Two horizontal trajectory strips side by side:
(a) Type~I example (HotpotQA episode) and (b) Type~D example
(APPS Interview episode).

\textbf{Data:} Select one representative episode from each environment
from deployment logs. Need per-step:
\begin{enumerate}[nosep]
\item Step index $t$
\item EAAG gate decision: trigger (1) or not (0)
\item Entropy value at that step
\item Whether the rollout actually helped ($U_t$)
\item What a fixed-direction baseline would have decided
\end{enumerate}

\textbf{Visual design (ref: CoRefine Figure~4 + custom):}
\begin{itemize}[nosep]
\item Each step is a colored circle on a horizontal timeline
\item Green circle = EAAG triggers and $U{=}1$ (correct trigger)
\item Gray circle = EAAG doesn't trigger (correct skip)
\item Red circle = fixed baseline triggers and $U{=}0$ (harmful trigger)
\item Circle size proportional to entropy level
\item Below each timeline: small entropy curve (line plot)
\item Above: annotations at key decision points
\end{itemize}

\textbf{Panel (a) HotpotQA (Type~I):} EAAG learns NOT to trigger
at high-entropy late steps (gray circles), while fixed baseline
triggers and gets harmed (red Xs overlaid).
\textbf{Panel (b) APPS Interview (Type~D):} EAAG triggers at
high-entropy steps (green circles), matching the fixed baseline
but with better selectivity.

\textbf{Key message:} Same high entropy $\to$ opposite gate decisions
$\to$ EAAG adapts correctly in both cases.
This is the ``Same Signal, Opposite Meaning'' story in one figure.

\textbf{Data source:} Select episodes from deployment logs where
the direction difference is most visible. Choose episodes where
fixed-direction baseline makes $\geq$2 wrong trigger decisions.}
