% ============================================================
% 4. METHOD: EAAG
% ============================================================
\section{Method: Environment-Aware Adaptive Gating}
\label{sec:method}

Proposition~\ref{prop:necessity} establishes that direction
discovery is necessary for cross-environment robustness; no
existing method performs it (Table~\ref{tab:classification}).
EAAG (Environment-Aware Adaptive Gating) closes this gap by
learning both signal identity and direction from online
interaction data.

\begin{table}[t]
\caption{Method classification by adaptive compute components. All
prior methods share \{Single signal, Fixed direction, Problem-level\}.
EAAG is the only method with \{Multi-signal, Learned direction,
Step-level, Environment-aware\}.}
\label{tab:classification}
\centering\small
\begin{tabular}{lcccc}
\toprule
Method & Signal & Direction & Granularity & Env-Aware \\
\midrule
CaTS$^\dagger$     & Single & Fixed & Problem & \xmark \\
SEAG$^\dagger$     & Single & Fixed & Problem & \xmark \\
CoRefine$^\dagger$ & Single & Fixed & Problem & \xmark \\
CATTS              & Single & Fixed & Problem & \xmark \\
AUQ                & Single & Fixed & Problem & \xmark \\
\midrule
\textbf{EAAG}      & \textbf{Multi (auto)} & \textbf{Learned}
                    & \textbf{Step} & \cmark \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Design rationale.}
The bottleneck in adaptive gating is direction, not model capacity
(\S\ref{sec:ablation}, Table~\ref{tab:capacity}): an MLP with the
wrong direction falls \emph{below} the no-trigger baseline, while
a logistic gate with the correct direction Pareto-dominates all
baselines. EAAG therefore pairs a rich discovery pipeline
(exploration + LLM reasoning + LASSO) with a deliberately simple
gate---a logistic function that encodes both signal identity and
direction.

{\color{red} \textbf{[FIGURE NEEDED: Figure 4.]} EAAG method overview diagram.
Three-phase pipeline: Explore (randomized gating, collect D) $\to$
Reason (LLM analyzes D, generates environment profile + feature hypotheses)
$\to$ Learn (LASSO selects features, trains logistic gate with signed weights).
Show arrows from \S3 findings (direction reversal, signal identity, necessity)
to each design choice.
Similar to FLARE's method section which has equations inline but
a clear conceptual flow. Consider a horizontal pipeline diagram.}

EAAG operates in three phases (Figure~\ref{fig:method}):
exploration, reasoning, and learning.

\subsection{Exploration}

The agent explores a new environment with $N_{\text{explore}} = 50$
episodes using randomized gating ($\varepsilon = 0.5$): at each step,
the optimizer $T$ is invoked with probability 0.5 regardless of state.
This produces a dataset $\mathcal{D} = \{(s_t^{(i)},\; U_t^{(i)})\}$
of state--utility pairs. The randomized design ensures unbiased
estimation of $U$ across the full signal range, a requirement that
targeted exploration would violate by conditioning on the signal
whose semantics are unknown.

\subsection{Feature Generation}

An LLM analyzes $\mathcal{D}$ to identify patterns distinguishing
useful from wasteful rollouts. It produces an \emph{environment
profile} describing the task structure and \emph{feature hypotheses}
proposing task-specific signals to extract. For example, in WebShop
the LLM identifies that rollouts help during product comparison but
not during search, proposing features such as
\texttt{price\_mentioned} and \texttt{action\_is\_click}. These
complement a set of universal features (step count, token entropy,
state length) available across all environments. The LLM's primary
value is generalizability: it enables zero-shot feature generation
for unseen environments where task-specific signals cannot be
anticipated by universal features alone.

\subsection{Sparse Direction Learning}

LASSO selects the most relevant features from the union of universal
and LLM-generated candidates, then trains a logistic regression gate:
\begin{equation}
  g(s) = \mathbf{1}\!\left[\sigma(\mathbf{w}^\top \phi(s) + b)
  > \tau\right]
  \label{eq:gate}
\end{equation}
where $\phi(s)$ is the selected feature vector and $\mathbf{w}$
encodes both direction (sign) and importance (magnitude) for each
signal. The sign of each $w_i$ directly implements direction
discovery: $w_i > 0$ triggers on high signal values (Type~D
semantics), $w_i < 0$ triggers on low values (Type~I semantics), and
$w_i = 0$ indicates an uninformative signal (LASSO selection).
Training completes in ${<}1$\,s on CPU. At deployment, the gate
adds zero per-step cost.

\subsection{Online Adaptation}

During deployment, EAAG optionally continues collecting data via
$\varepsilon$-greedy exploration ($\varepsilon: 0.1 \to 0$) and
retrains the gate periodically, enabling adaptation to distributional
shift.
